Efficient pedigree recording for fast population genetics simulation

### 2017-11-16 forwards_paper.tex
moved data format

\plr{Order is intro-results-discussion-methods.}

@@

@@

First we describe the data structures we use for recording genealogical history,
as implemented in \msprime.
These derive from those described by \citet{kelleher2016efficient},
but have been generalised and modified to remove redundancy.
The tables below give the example tree sequence of Figure \ref{fig:ex_tree_seq}.

\begin{figure}
    \begin{center}
        \includegraphics[width=0.6\textwidth]{example_tree_seq}
    \end{center}
    \caption{
        A pictorial representation of a tree sequence
        relating three samples to each other
        over a chromosome of length 1.0.
        Mutations shown in the example tables of the text
        are marked with red stars.
        \label{fig:ex_tree_seq}
    }
\end{figure}

To clarify terminology,
below a \emph{tree} refers to
a genealogical tree describing how a collection of individuals are related to each other.
A \emph{tree sequence} contains information sufficient to reconstruct the genealogical
tree relating all samples to each other at any point along the genome.
In the context of a tree sequence, \emph{nodes} refer to distinct ancestors,
and correspond to the vertices in the trees of a tree sequence.
Since each node represents a certain ancestor, it has a unique ``time'', thought of as her birth time,
which determines the height of any branching points she is associated with.  A given node will be
associated with branching points of all trees across a region if that node
is the most recent common ancestor to the subtending tips across that region.
This information is stored in the columns of a \textbf{Node Table}:
\begin{center}
\begin{tabular}{cccc}
    id   &  is\_sample &  population &  time \\
    0    &  1          & 0           & 0 \\
    1    &  1          & 0           & 0 \\
    2    &  1          & 0           & 0 \\
    3    &  0          & 0           & 0.3 \\
    4    &  0          & 0           & 0.5 \\
    5    &  0          & 0           & 0.6 \\
    6    &  0          & 0           & 0.7 \\
\end{tabular}
\end{center}
where ``flags'' records other information
(e.g., a binary mask of `1' indicates the node is a sample).
Importantly, the \textbf{node ID} of a node is given implicitly
by the (zero-based) index of its corresponding row in the Node Table.

Tree sequences are constructed by specifying over which segments of genome
which nodes inherit from which other nodes.  This information is stored by
recording the endpoints of each distinctly inherited ancestral segment,
the parental node, and a list of children nodes who have inherited that segment.
As each such record describes a collection of edges across a swatch of trees in the tree sequence,
we call these \emph{edgesets}
and store them in the columns of an \textbf{Edgeset Table:}
\begin{center}
\begin{tabular}{cccc}
    left & right & parent & children \\
    0.2  &  0.8  &  3     & 0,2 \\
    0.0  &  0.2  &  4     & 1,2 \\
    0.2  &  0.8  &  4     & 1,3 \\
    0.8  &  1.0  &  4     & 1,2 \\
    0.8  &  1.0  &  5     & 0,4 \\
    0.0  &  0.2  &  6     & 0,4  \\
\end{tabular}
\end{center}

To record information about genetic variants we need to also record each mutation
and which nodes have inherited that mutation.
The tree structure takes care of inheritance -- all we need to do
is to record the highest node in the tree at the mutated site
that inherited that mutation.
As more than one mtuation may occur at a given site,
we separate this information into two tables,
first, the \textbf{Site Table} records for each variant site
\begin{center}
\begin{tabular}{ccc}
    id  &   position  & ancestral\_state \\
    0   &  0.1        & 0 \\
    1   &  0.5        & 0 \\
\end{tabular}
\end{center}
Here ``position'' is a (floating point) position along the chromosome,
and ``ancestral state'' is the genotype of the root of the tree at that site.
As for nodes, \textbf{site IDs} are given implicitly
by the (zero-based) index of the rows.
Then, we record in a \textbf{Mutation Table}
\begin{center}
\begin{tabular}{ccc}
    site  & node  & derived\_state \\
    0	  & 4	  & 1 \\
    1	  & 3	  & 1 \\
    1	  & 2	  & 0 \\
\end{tabular}
\end{center}
in which ``site'' is the ID of the site at wihch this mutation occurred,
``node'' is the ID of the highest node that has inherited this mutation,
and ``derived state'' is the genotype at this site of any individuals inheriting this mutation,
unless another mutation occurs.
% \plr{Omit migrations?}

\paragraph{Definition of valid tables}
Here are the formal requirements for a set of nodes and edgesets to make sense,
and to allow ``msprime``'s algorithms to work properly.

To disallow time travel and multiple inheritance:
\begin{enumerate}
    \item Offspring must be born after their parents (and hence, no loops).
    \item The set of intervals on which each individual is a child must be disjoint.
\end{enumerate}
For algorithmic reasons, we also require:
\begin{enumerate} \setcounter{enumi}{2}
    \item The leftmost endpoint of each chromosome is 0.0.
    \item Node times must be strictly greater than zero.
    \item The list of offspring in an edgeset must be sorted.
    \item Edgesets must be sorted in nondecreasing time order.
    \item The set of intervals on which each individual is a parent must be disjoint.
    \item Each edgeset must contain at least two children.
\end{enumerate}
Note that since each node time is equal to the amount of time since the \emph{birth} of the
corresponding parent, time is measured in clock time, not in meioses.

A forwards-time simulation does \textbf{not} naturally emit genealogical information
satisfying requirements 5--8.
However, \msprime{} implements two algorithms that will take a set of tables
satisfying only 1--4 and produce tables satisfying all requirements.
Trivially, \texttt{sort\_tables} enforces requirements 5 and 6 and does not renumber nodes;
then, \texttt{simplify} enforces requirements 7 and 8 (and does much more; see below).

\section{More general method for recording the pedigree}

\plr{moved from above}

Concretely, this is done as follows.
Suppose that the forwards simulation algorithm labels (haploid) individuals by integers,
which we call ``input labels'',
to distinguish them from the ``node IDs'' given to these same individuals in the (output) tree sequence.
The algorithm maintains at all times a set of tables $(\nodes, \edgesets, \sites, \mutations)$
that record a relaxed tree sequence,
and an associative array $L$ that maps input labels to output node IDs, so that if $x$ is an input label,
then $L[x]$ is the corresponding output node ID.
We also always maintain $n$ to be the number of rows currently in the node table,
(so that with zero-indexed IDs, the next to be added will have node ID $n$),
$T_0$ to be the time of last simplification,
and $n_0$ the number of rows in the node table at that time.

Initially, we begin with $n$ and $n_0$ equal to the number of rows in the initial node table,
and $L[j] = i_j$ for each $0 \le j < N$ if the initial input generation is labeled
$i_0$, \ldots, $i_{N-1}$.

At a reproduction event where haploid parents $x$ and $y$ produce offspring $u$
at time $t$ in population $p$,
\begin{enumerate}
    \item add a
        $(
        \text{flags}=0,
        \text{population}=p,
        \text{time}=t)$ row to the node table,
    \item set $L[u] = n$,
    \item and increment $n\mathrel{+}=1$.
\end{enumerate}
Then, for each interval $[\ell,r)$ that $u$ inherits from parent $z$
(where $z$ is either $x$ or $y$),
\begin{enumerate}[resume]
    \item add a
        $( \text{left}=\ell,
        \text{right}=r,
        \text{parent}=z,
        \text{children}=(u,))$ row to the edgeset table.
\end{enumerate}
If furthermore there have been mutations at genomic locations $s_1$, \ldots, $s_k$
on this interval,
with derived states $q_1$, \ldots, $q_k$,
then for each $1 \le j \le k$,
\begin{enumerate}[resume]
    \item if $s_j$ is not in the site table, add a row
        $( \text{position}=s_j,
        \text{ancestral\_state}=0)$,
    \item find the site index $i$ whose position is $s_j$,
    \item and add a row
        $( \text{site}=i$,
           \text{node}=u$,
        \text{derived\_state}=q_j)$ to the mutation table.
\end{enumerate}

To \textbf{simplify} the tree sequence at time $T$,
\begin{enumerate}
    \item add $T-T_0$ to each time in the first $n_0$ rows of the node table,
        and replace each remaining time $t$ with $T-t$.
\end{enumerate}
Then, pass the set of currently alive input IDs,
$i_0, \ldots, i_{N-1}$ to the simplificaiton algorithm,
which produces a tree sequence that has node ID $j$ corresponding to input ID $i_j$
for $0 \le j < N$, and
\begin{enumerate}[resume]
    \item empty $L$,
    \item let $L[j] = i_j$ for $0 \le j < N$,
    \item set $n$ to be the number of nodes in the new tree sequence and set $n_0 = n$, and finally
    \item set $T_0 = T$.
\end{enumerate}

Simplification keeps the tables to a managable size.
Since the map $L$ is updated to maintain the association between individuals in the simulation
and nodes in the tree sequence, simplification can be run regularly,
as the simulation progresses.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}

>>>>

@@

Here we describe in more detail the data structures underlying a tree sequence in \msprime.
These derive from those described by \citet{kelleher2016efficient},
but have been generalised and modified to remove redundancy.

To clarify terminology,
below a \emph{tree} refers to
a genealogical tree describing how a collection of individuals are related to each other.
A \emph{tree sequence} contains information sufficient to reconstruct the genealogical
tree relating all samples to each other at any point along the genome.
In the context of a tree sequence, \emph{nodes} refer to distinct ancestors,
and correspond to the vertices in the trees of a tree sequence.
Since each node represents a certain ancestor, it has a unique ``time'', thought of as her birth time,
which determines the height of any branching points she is associated with.  A given node will be
associated with branching points of all trees across a region if that node
is the most recent common ancestor to the subtending tips across that region.
This information is stored in the columns of a \textbf{Node Table},
having columns \emph{time}, \emph{population}, and \emph{flags}.
The column ``flags'' records other information
(e.g., a binary mask of `1' indicates the node is a sample).
Importantly, the \textbf{node ID} of a node is given implicitly
by the (zero-based) index of its corresponding row in the Node Table.

Tree sequences are constructed by specifying over which segments of genome
which nodes inherit from which other nodes.  This information is stored by
recording the endpoints of each distinctly inherited ancestral segment,
the parental node, and the child node who inherited that segment.
As each such record describes particular edge across a swatch of trees in the tree sequence,
we call these \emph{edges}
and store them in the columns of an \textbf{Edge Table},
having columns \emph{left}, \emph{right}, \emph{parent}, and \emph{child}.

To record information about genetic variants we need to also record each mutation
and which nodes have inherited that mutation.
The tree structure takes care of inheritance -- all we need to do
is to record the highest node in the tree at the mutated site
that inherited that mutation.
As more than one mtuation may occur at a given site,
we separate this information into two tables,
first, the \textbf{Site Table} records for each variant site its \emph{position} and \emph{ancestral state}.
Here \emph{position} is a (floating point) position along the chromosome,
and \emph{ancestral state} is the genotype of the root of the tree at that site.
As for nodes, \textbf{site IDs} are given implicitly
by the (zero-based) index of the rows.
Then, we record in a \textbf{Mutation Table} the \emph{site}, \emph{node}, and \emph{derived state}
for each mutation.
Here ``site'' is the ID of the site at wihch this mutation occurred,
``node'' is the ID of the highest node that has inherited this mutation,
and ``derived state'' is the genotype at this site of any individuals inheriting this mutation,
unless another mutation occurs.
% \plr{Omit migrations?}

\paragraph{Definition of valid tables}
Here are the formal requirements for a set of nodes and edges to make sense,
and to allow ``msprime``'s algorithms to work properly.

To disallow time travel and multiple inheritance:
\begin{enumerate}
    \item Offspring must be born after their parents (and hence, no loops).
    \item The set of intervals on which each individual is a child must be disjoint.
\end{enumerate}
For algorithmic reasons, we also require:
\begin{enumerate} \setcounter{enumi}{2}
    \item The leftmost endpoint of each chromosome is 0.0.
    \item Node times must be strictly greater than zero.
    \item Edges must be sorted in nondecreasing time order.
\end{enumerate}
Note that since each node time is equal to the amount of time since the \emph{birth} of the
corresponding parent, time is measured in clock time, not in meioses.

A forwards-time simulation does \textbf{not} naturally emit genealogical information
satisfying all these requirements.
However, \msprime{} implements two algorithms that will take a set of tables
satisfying only 1--4 and produce tables satisfying all requirements.
These are \texttt{sort\_tables}, to enforce sortedness requirements (without renumbering nodes),
and \texttt{simplify}, which works on sorted tables
and is guarenteed to produce a valid, \msprime{}-ready tree sequence.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
Thanks to Brad Shaffer and Evan McCartney-Melstad for useful discussions.
Funding for this project cam from XXX
